Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Channel at the LHC with Multivariate Techniques

### 2015-11-30 paper/sec-intro.tex
work on the intro

As in the case of single Higgs production,
in the SM the dominant mechanism for the production of a pair of Higgs bosons at the LHC is 
gluon fusion (see Ref.~\cite{baglio} and references therein), analogously to single Higgs production~\cite{Dittmaier:2012vm}.
%
For a center-of-mass energy $\sqrt{s} = 14\,$TeV, the next-to-next-to-leading order (NNLO)
total cross section is approximately $40\,$fb~\cite{deFlorian:2013jea},
which is increased by a further few percent once NNLL corrections
are accounted for~\cite{deFlorian:2015moa}.
%
Feasibility studies in the case of a SM-like Higgs boson
in gluon fusion have been performed for several different final states, including
$b\bar b\gamma\gamma$~\cite{Baur:2003gp,Barger:2013jfa},
$b\bar{b}\tau^+\tau^-$~\cite{Baur:2003gpa,Barr:2013tda,Dolan:2012rv,Dolan:2013rja},
$b\bar{b}W^+W^-$~\cite{Dolan:2012rv,Papaefstathiou:2012qe} and
$b\bar{b}b\bar{b}$~\cite{Baur:2003gpa,Dolan:2012rv,Wardrope:2014kya,deLima:2014dta,Barger:2013jfa}.
%
While these studies differ in their qualitative conclusions,
a consistent picture emerges 
that the ultimate precision in the determination of the Higgs trilinear
coupling $\lambda$
will only be achieved from the combination of various final states.

@@

In the context of SM production,
the main advantage of the $b\bar{b}b\bar{b}$ final state is the
enhancement of the signal yield
from the large branching fraction of Higgs bosons into bottom-anti-bottom
pairs, ${\rm BR}\lp H\to b\bar{b}\rp\simeq 0.57$~\cite{Dittmaier:2012vm}.
%
However, a measurement in this channel
needs to deal with an overwhelming QCD multi-jet background.
%
Recent studies of Higgs pair production in this
channel~\cite{Wardrope:2014kya,deLima:2014dta}
estimate that with a luminosity of
$\mathcal{L}_{\rm int}$=3000 fb$^{-1}$ a signal significance of around $S/\sqrt{B}\sim 2.0$ could be obtained.
%
These studies are
based on exploiting the differences between signal and background
using techniques that account for a large number of kinematic
variables, Boosted Decision Trees
and Shower Reconstruction~\cite{Soper:2011cr} respectively.
%
A limitation of these works is that
they do not account
 for the light-jet background processes $bbjj$ and
$jjjj$, which can fake the signal when 
some of the light jets are mistagged as $b$-jets.
%
As will show, this is not a good approximation since both due to
combinatorics and to $b$-quark radiation in the shower, the
contribution of $2b2j$ and $4j$ can become comparable
to that of the irreducible $4b$ component.

In this work we revisit the feasibility of SM Higgs pair production by gluon-fusion
in the $4b$ final state.
%
 Our strategy is based on a combination of traditional cut-based
 methods with modern multivariate analysis (MVA)
 (in particular feed-forward multi-layer neural networks),
 which lead to a substantial
  enhancement of the signal discrimination.
  %
  We account for  all relevant
  backgrounds, including the contribution from light jet fakes.
  %
  By combining the contribution from the boosted, intermediate and resolved
  topologies we achieve a signal significance of $S/\sqrt{B}\simeq 3.2$.
%
  Therefore, our strategy allow us to improve the
  potential of this final state even when considering a
  more realistic background modeling.

  One of the original motivations of this study was to provide guidance to the
  LHC experiments
concerning which aspects of the measurement should be particularly optimized
to increase the signal significance in the $4b$ channel.
%
In this respect, we study the dependence on our results on the $
b$-tagging mistag
efficiency, the light-jet mistag rate and the 
momentum resolution.
%
In each of these cases, we determine the signal significance
$S/\sqrt{B}$ of the
measurement in both conservative and optimistic assumptions of these
analysis settings.

The structure of this paper is the following.
%
In Sect.~\ref{mcgeneration} we discuss the modeling of the signal
and background processes using Monte Carlo event generators.
%
In Sect.~\ref{sec:analysis}
we introduce our analysis strategy, and in particular
the classification of each individual event into
different categories according to their topology.
%
The results for the traditional cut-based analysis
are presented in Sect.~\ref{sec:results}, and in
Sect.~\ref{sec:mva} we show how we can enhance signal
significance using multivariate techniques.
%
The robustness of our results under realistic pile-up
conditions expected at the HL-LHC is illustrated in
 Sect.~\ref{sec:pileup}.
%
Finally in Sect.~\ref{sec:conclusions} we conclude and outline
how our results can be extended to estimate the accuracy
in the determination of $\lambda$ and the constraints on
BSM scenarios in an EFT framework.

>>>>

As in the case of single Higgs production,
in the SM the dominant mechanism for the production of a pair of Higgs bosons at the LHC is 
gluon fusion (see Ref.~\cite{baglio,Frederix:2014hta} and references therein), analogously to single
Higgs production~\cite{Dittmaier:2012vm}.
%
For a center-of-mass energy $\sqrt{s} = 14\,$TeV, the next-to-next-to-leading order (NNLO)
total cross section is approximately $40\,$fb~\cite{deFlorian:2013jea},
which is increased by a further few percent once NNLL corrections
are accounted for~\cite{deFlorian:2015moa}.
%
Feasibility studies in the case of a SM-like Higgs boson
in gluon fusion have been performed for several different final states, including
$b\bar b\gamma\gamma$~\cite{Baur:2003gp,Barger:2013jfa},
$b\bar{b}\tau^+\tau^-$~\cite{Baur:2003gpa,Barr:2013tda,Dolan:2012rv,Dolan:2013rja},
$b\bar{b}W^+W^-$~\cite{Dolan:2012rv,Papaefstathiou:2012qe} and
$b\bar{b}b\bar{b}$~\cite{Baur:2003gpa,Dolan:2012rv,Wardrope:2014kya,deLima:2014dta,Barger:2013jfa}.
%
While these studies differ in their qualitative conclusions,
a consistent picture emerges 
that the ultimate precision in the determination of the Higgs trilinear
coupling $\lambda$
will only be achieved from the combination of various final states.

@@

In the context of SM production,
the main advantage of the $b\bar{b}b\bar{b}$ final state is the
enhancement of the signal yield
from the large branching fraction of Higgs bosons into bottom-anti-bottom
pairs, ${\rm BR}\lp H\to b\bar{b}\rp\simeq 0.57$~\cite{Dittmaier:2012vm}.
%
However, a measurement in this channel
needs to deal with an overwhelming QCD multi-jet background.
%
Recent studies of Higgs pair production in this
final state~\cite{Wardrope:2014kya,deLima:2014dta}
estimate that, for an integrated
luminosity of
$\mathcal{L}_{\rm int}$=3000 fb$^{-1}$,
a signal significance of around $S/\sqrt{B}\sim 2.0$ could be obtained.
%
These studies are
based on exploiting the differences between signal and background
using techniques that account for a large number of kinematic
variables, Boosted Decision Trees
and Shower Deconstruction~\cite{Soper:2011cr} respectively.
%
While irreducible backgrounds, such as $4b$ and
$t\bar{t}$, are included, the contribution
from reducible processes, such as $bbjj$ and
$jjjj$, which can fake the signal when 
light jets are mistagged as $b$-jets, are not accounted for.
%
As will show, this approximation is not justified since, both due to
combinatorics and to $b$-quark radiation in the shower, the
contribution of the $2b2j$ and $4j$ backgrounds can become comparable or even larger
than the irreducible $4b$ component.

In this work, we revisit the feasibility of SM Higgs pair production by gluon-fusion
in the $b\bar{b}b\bar{b}$ final state.
%
 Our strategy is based on a combination of traditional cut-based
 methods with modern multivariate analysis (MVA),
 leading to a substantial
  enhancement of the signal discrimination.
  %
  We account for  all relevant
  backgrounds, including the contribution from light jet fakes.
  %
  We also model the impact of the pile-up (PU) effects expected at the
  HL-LHC by means of modern PU subtraction techniques.
  %
  By combining the contribution from the boosted, intermediate and resolved
  topologies we achieve a total signal significance of $S/\sqrt{B}\simeq 4.0$
  for $\mathcal{L}_{\rm int}$=3000 fb$^{-1}$.
%
  These results emphasize the crucial importance that the $b\bar{b}b\bar{b}$ final state
  has for studies of Higgs pair production and for the determination of
  $\lambda$, as well as the advantages of the use of modern statistical
  analysis methods.

The structure of this paper is the following.
%
In Sect.~\ref{mcgeneration} we discuss the modeling of the signal
and background processes using Monte Carlo event generators.
%
In Sect.~\ref{sec:analysis}
we introduce our analysis strategy, and in particular
the classification of each individual event into
different categories according to their topology.
%
The results for the traditional cut-based analysis
are presented in Sect.~\ref{sec:results}, and in
Sect.~\ref{sec:mva} we show how we can enhance signal
significance using multivariate techniques.
%
The robustness of our results under realistic pile-up
conditions expected at the HL-LHC is illustrated in
 Sect.~\ref{sec:pileup}.
%
Finally, in Sect.~\ref{sec:conclusions} we conclude and outline
how our results can be extended to estimate the accuracy
in the determination of trilinear coupling $\lambda$ and the constraints on
BSM scenarios.

### 2015-11-30 paper/sec-pileup.tex
work on the intro

\subsection{MVA revisited with PU effects}

After having discussed in some detail which are the effects
of PU in the various kinematic distributions that are used
as input to the MVA, we now revisit the analysis
of Sect.~\ref{sec:mva} now with PU included.

First of all we compare the distribution of the ANN output for the
signal and background MC events in the boosted and resolved categories,
the analogous comparison in the case without PU was shown
in Fig.~\ref{fig:nnresponse}.
%
As we can see in Fig.~\ref{fig:nnresponse_PU}, there is a clear discrimination
between signal and background in the two categories, also
in the case of PU included.
%
However, by comparing with Fig.~\ref{fig:nnresponse} we see
that some of the most distinguishing features that separate signal
and background in the ANN output are reduced when PU effects
are included.
%
For instance, in the boosted category, there is a trend of the distribution
for signal events to move towards smaller values of the
ANN discriminant.
%
This in turn will entail a reduction of the signal significance as
compared to the case without PU.

@@

Next we consider the number of signal and background events that
are expected at the HL-LHC as a function of the ANN discriminant -
the corresponding results without PU are those of
Fig.~\ref{fig:nev2}.
%
As we can see from Fig.~\ref{fig:nev2_PU}, the trends are
similar as in the no PU case, for instance, in the boosted
category we are left with around 200 events for the optimal value
of the ANN cut.
%
Comparing Figs.~\ref{fig:nev2} and~\ref{fig:nev2_PU}, we also observe
that the intermediate category is substantially degraded once PU
is included, since a reduced number of events fall into this
category now.

@@

Now we can finally revisit the post-MVA signal significance,
$S/sqrt{B}$, and he signal over background ratio, $S/B$,
in the case of PU, and compare with the corresponding
no PU results shown in Fig.~\ref{fig:sb_mva}.
%

In Fig.~\ref{fig:sb_mva_PU} we show the signal significance,
$S/\sqrt{B}$, as well as the signal over background ratio,
$S/B$, as a function of the NN discriminant, in the case
of the analysis including the effects of PU
with $\la n_{\rm PU}\ra=150$.
%
The corresponding results in the case without PU were shown in
Fig.~\ref{fig:sb_mva}.
%
As can be seen, the MVA-driven enhancement is robust event in the
presence of PU, and a total signal significance of
almost $S/\sqrt{B}\simeq 3$ is also obtained in this case.
%
Therefore, we conclude that the qualitative results obtained
in the previous section are robust even in the presence
of realistic PU effects.
%
Note that no specific effort has been performed to
optimize PU subtraction (for example by tuning the value
of the patch length $a$ in {\tt SoftKiller}), so there is
certainly still room for improvement.

@@

To conclude, it is illustrative to quantify which of the input variables
to the MVA are most significant in the case of PU,
and compare it with the corresponding
results without PU.
%
We use the same estimator as in Sect.~\ref{sec:signalsignificance},
namely the sum
of the absolute value of all the weights connected to a given
input neuron $i$, Eq.~(\ref{eq:totweight}).
%
The results for the resolved and boosted categories are shown
on Fig.~\ref{fig:nnweights_PU}.
%
As we can see by comparing to the corresponding
results without PU in Fig.~\ref{fig:nnweights}, we see that
in the boosted category, the discrimination power of the invariant
mass of Higgs candidates is decreased and that of the various substructure
variables, in particular $C_2$ and $D^{{\rm rm}}$, is conversely
increased.
%
This reflects the fact that these substructure variables are
relatively robust against PU contamination.

@@

\subsection{Combined results}

Finally, we now combine the three categories, resolved,
intermediate and boosted, and determine the signal
significance taking into account all event topologies.
%
First of all, in Table~\ref{table:cutflowMVA_PU}
we provide the  the number of signal and
    background events at the HL-LHC, as well as the $S/\sqrt{B}$ and $S/B$,
    for both the cut-based analysis (corresponding
    to $y_{\rm cut}=0$) and after the
    optimal MVA cut.
    %
    These numbers corresponding to the simulations
    with $\la n_{\rm PU}\ra=80$
    and SK subtraction.
    %
    The corresponding results without PU were collected
    in Table.~\ref{table:cutflowMVA}.
    %
    The value of the optimal MVA cut $y_{\rm cut}$ is determined
    as the one leading to highest signal significance, with
    the constraint that number of MC events left is still large
    enough to tame statistical fluctuations.

From Table~\ref{table:cutflowMVA_PU} we see that, thanks
to the MVA, we can improve the signal significance and the
signal over background ratio from 0.31 and $3\cdot 10^{-4}$
(0.40 and $1.3\cdot 10^{-4}$) up to 2.9 and 0.034 (2.5 and 0.15)
in the boosted (resolved) category.
%
The intermediate category exhibits a lower value of $S/\sqrt{B}$
for the optimal $y_{\rm cut}$ value, around unity.
%
We thus conclude that the boosted category is the most promising
one for the study of Higgs pair production in the $b\bar{b}b\bar{b}$
at the HL-LHC, with however useful information also provided
by the resolved category.

Adding in quadrature the signal significance from the three
(exclusive) categories, our final combined result is
\be
\lp \frac{S}{\sqrt{B}}\rp_{\rm tot} \simeq 4.0 \, ,
\ee
far above the threshold for {\it evidence}, and close to the
requirement for {\it discovery}.
%
In addition, it should be emphasized that there is still
room for improvement, in particular by means of a tailored
study dedicated to optimize PU subtraction in this process.

In summary, in this section we have demonstrated the robustness
of our results under the presence of PU conditions as those
expected at the HL-LHC.
%
With a signal significance of $S/\sqrt{B}\sim 4$, is clear that
the $b\bar{b}b\bar{b}$ final state can be used to claim evidence
for Higgs pair production at the HL-LHC without the need
to combine with any other final state.
%
It will be important now to quantify the precision of
a extraction of the Higgs trilinear coupling $\lambda$ using
the techniques presented here, but this is left for future work.

>>>>

\subsection{MVA revisited in the presence of PU}

After this discussion of the effects 
of PU in the various kinematic distributions
that are used
as input to the MVA, we now revisit the analysis
of Sect.~\ref{sec:mva} this time with PU included.
%
First of all we compare in Fig.~\ref{fig:nnresponse_PU}
the distribution of the ANN output for the
signal and background MC events in the boosted and resolved categories --
the analogous comparison in the case without PU was shown
in Fig.~\ref{fig:nnresponse}.
%
We observe that the clear MVA discrimination
between signal and background events in the two categories is robust
when PU is included.
%
However, by comparing with Fig.~\ref{fig:nnresponse}, we see
that some of the most distinguishing features that separate signal
and background in the ANN output are reduced when PU effects
are included.
%
For instance, in the boosted category, there is a trend of the distribution
for signal events to move towards smaller values of the
ANN discriminant.
%
This in turn will entail a reduction of the signal significance as
compared to the case without PU.

@@

In Table~\ref{table:cutflowMVA_PU}
we provide the  the number of signal and
    background events at the HL-LHC, as well as the $S/\sqrt{B}$ and $S/B$,
    for both the cut-based analysis (corresponding
    to $y_{\rm cut}=0$) and after the
    optimal MVA cut.
    %
    These numbers corresponding to the simulations
    with $\la n_{\rm PU}\ra=80$
    and SK subtraction.
    %
    The corresponding results without PU were collected
    in Table.~\ref{table:cutflowMVA}.
    %
    From Table~\ref{table:cutflowMVA_PU} we see that, thanks
to the MVA, we can improve the signal significance and the
signal over background ratio from 0.31 and $3\cdot 10^{-4}$
(0.40 and $1.3\cdot 10^{-4}$) up to 2.9 and 0.034 (2.5 and 0.15)
in the boosted (resolved) category.
%
The intermediate category exhibits a lower value of $S/\sqrt{B}$
for the optimal $y_{\rm cut}$ value, around unity.
%
We thus conclude that the boosted category is the most promising
one for the study of Higgs pair production in the $b\bar{b}b\bar{b}$
at the HL-LHC, with however useful information also provided
by the resolved category.

Next we show the results
as a function of the optimal ANN cut $y_{\rm cut}$.
%
In Fig.~\ref{fig:nev2_PU}
we show the number of signal and background events that
are expected at the HL-LHC as a function of the ANN discriminant -
the corresponding results without PU are those of
Fig.~\ref{fig:nev2}.
%
As we can see from Fig.~\ref{fig:nev2_PU}, the trends are
similar as those in the case without PU.
%
Comparing Figs.~\ref{fig:nev2} and~\ref{fig:nev2_PU}, we also observe
that the intermediate category is substantially degraded once PU effects
are accounted for, the reason being a reduction in the number of signal
events that are now assigned to this category.
%
This shows that the definition of the intermediate category is seriously
affected by PU, as opposed to the resolved and boosted selections.

@@

In Fig.~\ref{fig:sb_mva_PU} we show the signal significance,
$S/\sqrt{B}$, as well as the signal over background ratio,
$S/B$, as a function of the ANN discriminant $y_{\rm cut}$, in the case
of the analysis including the effects of PU
with $\la n_{\rm PU}\ra=80$.
%
The corresponding results in the case without PU were shown in
Fig.~\ref{fig:sb_mva}.
%
As can be seen, the MVA-driven enhancement is robust in the
presence of PU.
%
While unsurprisingly there us a degradation as compared to the case wo PU,
we still manage to achieve a signal significance of
around $S/\sqrt{B}\simeq 2.5$ separately for the boosted and resolved
categories.
%
On the other hand, the intermediate category becomes irrelevant.
%
We conclude that the qualitative conclusions drawn in
Sect.~\ref{sec:mva} are robust in the presence
of realistic PU effects.
%
Note that no specific effort has been performed to
optimize PU subtraction, for example by tuning the value
of the patch length $a$ in {\tt SoftKiller}, so there is
still room for improvement in this respect.

@@

It is illustrative to quantify which of the input variables
to the MVA are most significant in the case of PU,
and compare it with the corresponding
results without PU.
%
We use the same estimator as in Sect.~\ref{sec:signalsignificance},
namely the sum
of the absolute value of all the weights connected to a given
input neuron $i$, Eq.~(\ref{eq:totweight}).
%
The results for the resolved and boosted categories are shown
on Fig.~\ref{fig:nnweights_PU}.
%
As we can see by comparing to the corresponding
results without PU in Fig.~\ref{fig:nnweights}, we see that
in the boosted category, the discrimination power of the invariant
mass of Higgs candidates is decreased and that of the various substructure
variables, in particular $C_2^{\beta}$ and
$D^{{\beta}}$, is conversely
increased.
%
This reflects the fact that these substructure variables are
relatively robust against PU contamination.

@@

Finally, we now combine the three categories, resolved,
intermediate and boosted, and determine the signal
significance taking into account all event topologies.
%
Adding in quadrature the signal significance from the three
(exclusive) categories for the corresponding
optimal value of $y_{\rm cut}$, our final combined result is
\be
\lp \frac{S}{\sqrt{B}}\rp_{\rm tot} \simeq 4.0 \, ,
\ee
that is, well
above the threshold for evidence, and  relatively close to the
requirement for discovery.
%

In summary, in this section we have demonstrated the robustness
of our results under the presence of PU conditions as those
expected at the HL-LHC.
%
With a signal significance of $S/\sqrt{B}\sim 4$, is clear that
the $b\bar{b}b\bar{b}$ final state can be used to claim evidence
for Higgs pair production at the HL-LHC without the need
to combine with any other final state.
%
Our analysis certainly contains
room for improvement, in particular by means of a tailored
study dedicated to optimize PU subtraction in this process.
%
It will be important now to quantify the precision of
a extraction of the Higgs trilinear coupling $\lambda$ using
the techniques presented here, but this is left for future work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
